### 2024-11-27 paper/paper.qmd
EDA update

---
title: "Analysis of Shootings and Firearm Discharges in Toronto"
subtitle: "A Comprehensive Examination Reveals Significant Time and Locational Impacts on Incident Severity"
author: 
  - Xinze Wu
thanks: "Code and data are available at: https://github.com/ke3w/Data-Analysis-of-Shootings-and-Firearm-Discharges-in-Toronto."
date: 2024/11/26
date-format: long
abstract: "This paper analyzes the impact of time and location on the severity of shootings and firearm discharges in Toronto. Utilizing a Bayesian generalized linear model, significant variations across neighborhoods and time ranges are revealed, emphasizing the need for targeted policy interventions. This study not only aids in understanding the dynamics of urban violence but also assists in resource allocation for law enforcement and public safety initiatives."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(palmerpenguins)
```

# Introduction

Overview paragraph

Estimand paragraph

Results paragraph

Why it matters paragraph

Telegraphing paragraph: The remainder of this paper is structured as follows. @sec-data....

# Data {#sec-data}

## Overview

We use the statistical programming language R [@citeR].... Our data [@shelter].... Following @tellingstories, we consider...

Overview text

## Measurement

Some paragraphs about how we go from a phenomena in the world to an entry in the dataset.

## Outcome variables

Add graphs, tables and text. Use sub-sub-headings for each outcome variable or update the subheading to be singular.

Some of our data is of penguins (@fig-bills), from @palmerpenguins.

```{r}
#| label: fig-bills
#| fig-cap: Bills of penguins
#| echo: false

ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange","purple","cyan4"),
                    guide = "none") +
  theme_minimal() +
  facet_wrap(~species, ncol = 1) +
  coord_flip()
```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

analysis_data <- read_csv(here::here("data/02-analysis_data/analysis_data.csv"))

analysis_data |> 
  ggplot(aes(x = width, y = length)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "Wing width (mm)",
       y = "Wing length (mm)")
```

Talk way more about it.

## Predictor variables

Add graphs, tables and text.

Use sub-sub-headings for each outcome variable and feel free to combine a few into one if they go together naturally.

# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.

### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

Please don't use these as sub-heading labels - change them to be what your point actually is.

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}

# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")
```

\newpage

# References

### 2024-11-27 scripts/05-exploratory_data_analysis.R

@@ -36,6 +36,9 @@ analysis_data$division <- factor(analysis_data$division, levels = reorder_factor
 analysis_data$occ_time_range <- factor(analysis_data$occ_time_range, levels = reorder_factors(analysis_data, occ_time_range))
 analysis_data$neighbourhood_158 <- factor(analysis_data$neighbourhood_158, levels = reorder_factors(analysis_data, neighbourhood_158))
 
+# Convert 'occ_hour' to numeric
+analysis_data$occ_hour <- as.numeric(as.character(analysis_data$occ_hour))
+
 #### Plotting ####
 
 # Relationships by Date
@@ -78,7 +81,7 @@ ggplot(analysis_data, aes(x = division, y = weighted_score, fill = division)) +
 #### Model data ####
 # Predicting the weighted score based on various factors
 score_model <- stan_glm(
-  formula = weighted_score ~ occ_date + occ_doy + occ_hour + neighbourhood_158 + division,
+  formula = weighted_score ~ occ_date + occ_doy + occ_time_range + neighbourhood_158 + division,
   data = analysis_data,
   family = gaussian(), 
   prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
@@ -86,5 +89,7 @@ score_model <- stan_glm(
   seed = 853
 )
 
-#### Save model ####
-saveRDS(score_model, file = "models/first_model.rds")
+summary(score_model)
+
+# Save model
+saveRDS(model_validation_train, file = "models/fisrt_model.rds")

### 2024-11-28 scripts/05-exploratory_data_analysis.R
update predictors

@@ -35,9 +35,8 @@ reorder_factors <- function(data, grouping_var) {
 analysis_data$division <- factor(analysis_data$division, levels = reorder_factors(analysis_data, division))
 analysis_data$occ_time_range <- factor(analysis_data$occ_time_range, levels = reorder_factors(analysis_data, occ_time_range))
 analysis_data$neighbourhood_158 <- factor(analysis_data$neighbourhood_158, levels = reorder_factors(analysis_data, neighbourhood_158))
+analysis_data$occ_hour <- factor(analysis_data$occ_hour, levels = reorder_factors(analysis_data, occ_hour))
 
-# Convert 'occ_hour' to numeric
-analysis_data$occ_hour <- as.numeric(as.character(analysis_data$occ_hour))
 
 #### Plotting ####
 
@@ -81,7 +80,7 @@ ggplot(analysis_data, aes(x = division, y = weighted_score, fill = division)) +
 #### Model data ####
 # Predicting the weighted score based on various factors
 score_model <- stan_glm(
-  formula = weighted_score ~ occ_date + occ_doy + occ_time_range + neighbourhood_158 + division,
+  formula = weighted_score ~ occ_date + occ_doy + occ_hour + neighbourhood_158 + division,
   data = analysis_data,
   family = gaussian(), 
   prior = normal(location = 0, scale = 2.5, autoscale = TRUE),

### 2024-11-28 scripts/03-clean_data.R
model change

@@ -9,7 +9,7 @@
 #### Workspace setup ####
 library(tidyverse)
 library(janitor)
-library(lubridate)  # Ensure lubridate is loaded for date manipulation
+library(lubridate)
 
 #### Clean data ####
 # Load the raw shootings data
@@ -23,7 +23,8 @@ cleaned_data <-
     occ_date = ymd(occ_date),  # Convert occ_date to date format
     division = as.factor(division),  # Convert police division to factor
     death = as.integer(death),  # Ensure death count is integer
-    injuries = as.integer(injuries)  # Ensure injuries count is integer
+    injuries = as.integer(injuries),  # Ensure injuries count is integer
+    occ_dow = as.factor(occ_dow)
   ) %>%
   select(-hood_140, -neighbourhood_140, -long_wgs84, -lat_wgs84, -geometry) %>%
   drop_na()  # Optional: Drop rows with any NA values

### 2024-11-28 scripts/05-exploratory_data_analysis.R

@@ -21,23 +21,6 @@ analysis_data <- analysis_data %>%
   mutate(weighted_score = death * 2 + injuries)  # Adjust the greater importance of 'death' over 'injuries'
 
 #### Exploratory Data Analysis ####
-
-# Prepare a function to reorder factors based on the weighted score
-reorder_factors <- function(data, grouping_var) {
-  data %>%
-    group_by({{ grouping_var }}) %>%
-    summarise(avg_score = mean(weighted_score, na.rm = TRUE)) %>%
-    arrange(avg_score) %>%
-    pull({{ grouping_var }})
-}
-
-# Apply the function to various factors
-analysis_data$division <- factor(analysis_data$division, levels = reorder_factors(analysis_data, division))
-analysis_data$occ_time_range <- factor(analysis_data$occ_time_range, levels = reorder_factors(analysis_data, occ_time_range))
-analysis_data$neighbourhood_158 <- factor(analysis_data$neighbourhood_158, levels = reorder_factors(analysis_data, neighbourhood_158))
-analysis_data$occ_hour <- factor(analysis_data$occ_hour, levels = reorder_factors(analysis_data, occ_hour))
-
-
 #### Plotting ####
 
 # Relationships by Date
@@ -80,7 +63,7 @@ ggplot(analysis_data, aes(x = division, y = weighted_score, fill = division)) +
 #### Model data ####
 # Predicting the weighted score based on various factors
 score_model <- stan_glm(
-  formula = weighted_score ~ occ_date + occ_doy + occ_hour + neighbourhood_158 + division,
+  formula = weighted_score ~ occ_date + occ_doy + occ_time_range + neighbourhood_158 + division,
   data = analysis_data,
   family = gaussian(), 
   prior = normal(location = 0, scale = 2.5, autoscale = TRUE),

### 2024-11-28 scripts/03-clean_data.R
cleandata update

@@ -24,7 +24,6 @@ cleaned_data <-
     division = as.factor(division),  # Convert police division to factor
     death = as.integer(death),  # Ensure death count is integer
     injuries = as.integer(injuries),  # Ensure injuries count is integer
-    occ_dow = as.factor(occ_dow)
   ) %>%
   select(-hood_140, -neighbourhood_140, -long_wgs84, -lat_wgs84, -geometry) %>%
   drop_na()  # Optional: Drop rows with any NA values

### 2024-11-28 scripts/05-exploratory_data_analysis.R
model update

@@ -8,8 +8,8 @@
 
 #### Workspace setup ####
 library(tidyverse)
-library(rstanarm)
-library(dplyr)
+library(gbm)
+library(caret)
 library(ggplot2)
 
 #### Read data ####
@@ -30,6 +30,13 @@ ggplot(analysis_data, aes(x = occ_date, y = weighted_score)) +
   labs(title = "Weighted Score Over Time", x = "Date", y = "Weighted Score") +
   theme_minimal() +
   theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
+
+# Relationships by Day of Week
+ggplot(analysis_data, aes(x = occ_dow, y = weighted_score)) +
+  geom_bar(stat = "identity", fill = "steelblue") +
+  labs(title = "Weighted Score by Day of Week", x = "Day of Week", y = "Weighted Score") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
 
 # Relationships by Day of Year
 ggplot(analysis_data, aes(x = occ_doy, y = weighted_score)) +
@@ -60,18 +67,25 @@ ggplot(analysis_data, aes(x = division, y = weighted_score, fill = division)) +
   theme_minimal() +
   theme(axis.text.x = element_text(angle = 90, hjust = 1))
 
-#### Model data ####
-# Predicting the weighted score based on various factors
-score_model <- stan_glm(
-  formula = weighted_score ~ occ_date + occ_doy + occ_time_range + neighbourhood_158 + division,
-  data = analysis_data,
-  family = gaussian(), 
-  prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
-  prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
-  seed = 853
-)
+analysis_data$occ_date <- as.numeric(analysis_data$occ_date)
+analysis_data$occ_dow <- as.factor(analysis_data$occ_dow)
+analysis_data$occ_time_range <- as.factor(analysis_data$occ_time_range)
+analysis_data$neighbourhood_158 <- as.factor(analysis_data$neighbourhood_158)
+analysis_data$division <- as.factor(analysis_data$division)
 
-summary(score_model)
+### Model setup
+set.seed(123)  # for reproducibility
+model <- gbm(weighted_score ~ occ_date + occ_dow + occ_doy + occ_time_range + neighbourhood_158 + division,
+                 data = analysis_data,
+                 distribution = "gaussian",
[…]

### 2024-11-28 scripts/06-model.R

@@ -1,0 +1,58 @@
+#### Preamble ####
+# Purpose: Validate and check the model built on Toronto shootings data
+# Author: Xinze Wu
+# Date: 2024/11/28
+# Contact: kerwin.wu@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: tidyverse, rstanarm, caret
+
+#### Workspace setup ####
+library(tidyverse)
+library(rstanarm)
+library(caret)
+
+#### Load data ####
+analysis_data <- read_csv("data/02-analysis_data/analysis_data.csv")
+
+#### Prepare Data ####
+# Creating a weighted score for deaths and injuries
+analysis_data <- analysis_data %>%
+  mutate(weighted_score = death * 2 + injuries)
+
+#### Split Data into Training and Testing Sets ####
+set.seed(123)  # Ensure reproducibility
+split <- createDataPartition(analysis_data$weighted_score, p = 0.7, list = FALSE)
+training_data <- analysis_data[split, ]
+testing_data <- analysis_data[-split, ]
+
+#### Model Fitting ####
+# Fit the model on training data
+model_validation_train <- stan_glm(
+  formula = weighted_score ~ occ_date + occ_doy + occ_time_range + neighbourhood_158 + division,
+  data = training_data,
+  family = gaussian(),
+  prior = normal(0, 2.5, autoscale = TRUE),
+  prior_intercept = normal(0, 2.5, autoscale = TRUE),
+  seed = 853
+)
+
+#### Posterior Predictive Checks ####
+# Performing posterior predictive checks to assess how well the model predictions align with the actual data.
+pp_check(model_validation_train)
+
+#### Predictions on Test Data ####
+# Predicting on testing data
+testing_data <- testing_data %>%
+  mutate(predicted_scores = posterior_predict(model_validation_train, newdata = testing_data, type = "response") |> colMeans())
+
+#### Model Performance Evaluation ####
+# Calculate Root Mean Squared Error (RMSE) and R-squared to quantify prediction accuracy.
+rmse_value <- rmse(testing_data$weighted_score, testing_data$predicted_scores)
+r_squared_value <- rsquared(testing_data$weighted_score, testing_data$predicted_scores)
+
+# Print performance metrics
+cat("RMSE:", rmse_value, "\n")
+cat("R-squared:", r_squared_value, "\n")
+
+#### Save model for future use ####
+saveRDS(model_validation_train, file = "models/model_validation_train.rds")

### 2024-11-28 scripts/06-model.R
model analysis update

@@ -1,58 +1,55 @@
 #### Preamble ####
-# Purpose: Validate and check the model built on Toronto shootings data
+# Purpose: Load and further analyze the Gradient Boosting Machine model for predicting weighted scores of shootings and firearm discharges in Toronto.
 # Author: Xinze Wu
-# Date: 2024/11/28
+# Date: 2024/11/24
 # Contact: kerwin.wu@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: tidyverse, rstanarm, caret
+# Pre-requisites: The `tidyverse`, `gbm`, `caret` packages must be installed and loaded
 
 #### Workspace setup ####
 library(tidyverse)
-library(rstanarm)
+library(gbm)
 library(caret)
 
+#### Load saved model ####
+model <- readRDS("models/first_model.rds")
+
 #### Load data ####
+# Load the cleaned and preprocessed data for further analysis or prediction
 analysis_data <- read_csv("data/02-analysis_data/analysis_data.csv")
-
-#### Prepare Data ####
-# Creating a weighted score for deaths and injuries
+analysis_data$occ_date <- as.numeric(analysis_data$occ_date)
+analysis_data$occ_dow <- as.factor(analysis_data$occ_dow)
+analysis_data$occ_time_range <- as.factor(analysis_data$occ_time_range)
+analysis_data$neighbourhood_158 <- as.factor(analysis_data$neighbourhood_158)
+analysis_data$division <- as.factor(analysis_data$division)
 analysis_data <- analysis_data %>%
-  mutate(weighted_score = death * 2 + injuries)
+  mutate(weighted_score = death * 2 + injuries)  # Adjust the greater importance of 'death' over 'injuries'
 
-#### Split Data into Training and Testing Sets ####
-set.seed(123)  # Ensure reproducibility
-split <- createDataPartition(analysis_data$weighted_score, p = 0.7, list = FALSE)
-training_data <- analysis_data[split, ]
-testing_data <- analysis_data[-split, ]
 
-#### Model Fitting ####
-# Fit the model on training data
-model_validation_train <- stan_glm(
-  formula = weighted_score ~ occ_date + occ_doy + occ_time_range + neighbourhood_158 + division,
-  data = training_data,
-  family = gaussian(),
-  prior = normal(0, 2.5, autoscale = TRUE),
-  prior_intercept = normal(0, 2.5, autoscale = TRUE),
-  seed = 853
-)
[…]

### 2024-11-28 scripts/06-model.R
update model

@@ -46,7 +46,7 @@ predictions <- predict(model, newdata = analysis_data, n.trees = 500, type = "re
 actual <- analysis_data$weighted_score
 rmse <- sqrt(mean((predictions - actual)^2))
 
-#### Save performance metrics ####
+#### performance metrics ####
 # Show RMSE
 print(sprintf("RMSE: %f", rmse))
 

### 2024-11-28 scripts/05-exploratory_data_analysis.R
ModelUpdate

@@ -86,6 +86,7 @@ model <- gbm(weighted_score ~ occ_date + occ_dow + occ_doy + occ_time_range + ne
 
 # Model summary
 print(model)
+summary(model, n.trees = 500, plotit = TRUE)
 
 # Save model
 saveRDS(model, file = "models/first_model.rds")

### 2024-11-28 scripts/06-model.R

@@ -39,9 +39,6 @@ summary(model, n.trees = 500, plotit = TRUE)
 # Reset to default (optional, to avoid affecting other plots)
 par(cex.axis = 1)
 
-# Generate predictions on new or existing data
-predictions <- predict(model, newdata = analysis_data, n.trees = 500, type = "response")
-
 # Compare predictions with actual data and calculate performance metrics
 actual <- analysis_data$weighted_score
 rmse <- sqrt(mean((predictions - actual)^2))
@@ -49,7 +46,3 @@ rmse <- sqrt(mean((predictions - actual)^2))
 #### performance metrics ####
 # Show RMSE
 print(sprintf("RMSE: %f", rmse))
-
-#### Documentation ####
-# Document insights and potential next steps
-write("Detailed analysis of the model's performance and recommended improvements are documented.", file = "docs/model_insights.txt")

### 2024-11-29 paper/paper.qmd
paper update

@@
 author:
-  - Xinze Wu
+  - name: Xinze Wu
+    affiliation: University of Toronto
+    email: kerwin.wu@mail.utoronto.ca
 thanks: "Code and data are available at: https://github.com/ke3w/Data-Analysis-of-Shootings-and-Firearm-Discharges-in-Toronto."
-date: 2024/11/26
-date-format: long
-abstract: "This paper analyzes the impact of time and location on the severity of shootings and firearm discharges in Toronto.
-Utilizing a Bayesian generalized linear model, significant variations across neighborhoods and time ranges are revealed, emphasizing the need for targeted policy interventions.
-This study not only aids in understanding the dynamics of urban violence but also assists in resource allocation for law enforcement and public safety initiatives."
-format: pdf
-number-sections: true
-bibliography: references.bib
+date: "2024-11-26"
+output: 
+  pdf_document:
+    toc: true
+    number_sections: true
+abstract: |
+  This paper analyzes the impact of time and location on the severity of shootings and firearm discharges in Toronto, utilizing a Gradient Boosting Machine model.
+The findings highlight significant variations across different times and locations, emphasizing the need for targeted policy interventions.
+This study not only aids in understanding the dynamics of urban violence but also assists in resource allocation for law enforcement and public safety initiatives.
+bibliography: [references.bib]
 ---
-
-```{r}
-#| include: false
-#| warning: false
-#| message: false
-
-library(tidyverse)
-library(palmerpenguins)
-```
 
@@
 
-Overview paragraph
+The safety and security of urban areas are of paramount importance to their residents and governance structures.
+In Toronto, the occurrence of shootings and firearm discharges poses significant challenges to public safety.
+Understanding the patterns and determinants of these violent events is crucial for developing effective interventions and policies.
+This paper employs a Gradient Boosting Machine model to analyze how time-related factors and geographic locations influence the severity of shooting incidents in Toronto.
 
-Estimand paragraph
+## Estimand
 
-Results paragraph
+This study aims to estimate the effect of time (day of week, hour, and time range) and location (neighbourhood and police division) on the severity of shootings and firearm discharges in Toronto, measured by a weighted score combining deaths and injuries.
 
-Why it matters paragraph
+## Model Setup
 
-Telegraphing paragraph: The remainder of this paper is structured as follows. @sec-data....
+We utilize a Gradient Boosting Machine model, chosen for its robustness in handling various types of data and its ability to model complex interactions and nonlinear relationships between predictors and the response variable.
+
+## Importance of the Study
+
+The results of this study are crucial for informing public safety strategies and police resource allocation in Toronto, aiming to reduce the incidence and severity of violent firearm-related incidents.
 
@@
 
-## Overview
+Data for this study is sourced from Open Data Toronto: <https://open.toronto.ca/dataset/shootings-firearm-discharges/>, detailing all recorded shootings and firearm discharges within the city limits from 2004 to 2024.
 
-We use the statistical programming language R [@citeR].... Our data [@shelter].... Following @tellingstories, we consider...
+## Data Cleaning
 
-Overview text
+Data cleaning was performed using R[@citeR] packages such as `tidyverse` and `lubridate`.
+The `occ_date` variable was converted to a numeric format to fit the model requirements, and categorical variables were transformed using appropriate factor conversions.
 
-## Measurement
+## Variables
 
-Some paragraphs about how we go from a phenomena in the world to an entry in the dataset.
+### Outcome Variable
 
-## Outcome variables
+-   **Weighted Score**: Combines the number of deaths and injuries, weighting deaths twice as heavily(weighted_score = death \* 2 + injuries).
 
-Add graphs, tables and text. Use sub-sub-headings for each outcome variable or update the subheading to be singular.
+### Predictor Variables
 
-Some of our data is of penguins (@fig-bills), from @palmerpenguins.
+-   **Time Factors**:
+
+-   `occ_date`(date): Date of Offence Occurred
+
+-   `occ_dow` (day of the week): Day of the Month Offence Occurred
+
+-   `occ_time_range` (morning, afternoon, evening):Time Range of Day Offence Occurred
+
+-   **Location Factors**:
+
+-   `neighbourhood_158:`Name of Neighborhood using City of Toronto's new 158 neighborhood structure
+
+-   `division`: Police division where offence occurred
+
+## Data visualization
+
+### Weighted Score Over Time
 
 ```{r}
-#| label: fig-bills
-#| fig-cap: Bills of penguins
+#| warning: false
+#| message: false
+#| label: fig-date
+#| fig_caption: true
+#| fig-cap: "line graph shows the relationship between weighted score over date"
 #| echo: false
 
-ggplot(penguins, aes(x = island, fill = species)) +
-  geom_bar(alpha = 0.8) +
-  scale_fill_manual(values = c("darkorange","purple","cyan4"),
-                    guide = "none") +
+ggplot(analysis_data, aes(x = occ_date, y = weighted_score)) +
+  geom_line(group = 1, color = "blue") +
+  scale_x_date(date_breaks = "1 year", date_labels = "%Y") +
+  labs(title = "Weighted Score Over Time", x = "Date", y = "Weighted Score") +
   theme_minimal() +
-  facet_wrap(~species, ncol = 1) +
-  coord_flip()
+  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
 ```
 
-Talk more about it.
+This plot @fig-date illustrates the variation in the weighted scores of shootings and firearm discharges in Toronto from 2004 to 2025.
+The weighted score, which combines the number of deaths and injuries with deaths being given a higher weight, is plotted against the date of incidents.
+Notably, the plot reveals several significant spikes, indicating periods with higher shooting severity.
+These peaks might correlate with specific events or changes in local circumstances, warranting further investigation.
+The majority of the time, however, the weighted scores remain relatively low, suggesting sporadic rather than consistent patterns of high-severity incidents.
+This visualization highlights the dynamic nature of crime severity over time and underscores the importance of continuous monitoring and analysis to understand the underlying trends and triggers.
 
-And also planes (@fig-planes).
-(You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)
+### Weighted Score by Day of week
 
 ```{r}
-#| label: fig-planes
-#| fig-cap: Relationship between wing length and width
-#| echo: false
 #| warning: false
 #| message: false
+#| label: fig-dow
+#| fig_caption: true
+#| fig-cap: "Bar chart of the weighted score by the day of week"
+#| echo: false
 
-analysis_data <- read_csv(here::here("data/02-analysis_data/analysis_data.csv"))
-
-analysis_data |> 
-  ggplot(aes(x = width, y = length)) +
-  geom_point(alpha = 0.8) +
+ggplot(analysis_data, aes(x = occ_dow, y = weighted_score)) +
+  geom_bar(stat = "identity", fill = "steelblue") +
+  labs(title = "Weighted Score by Day of Week", x = "Day of Week", y = "Weighted Score") +
   theme_minimal() +
-  labs(x = "Wing width (mm)",
-       y = "Wing length (mm)")
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
 ```
 
-Talk way more about it.
+The bar chart @fig-dow illustrates the weighted scores of shootings and firearm discharges in Toronto, broken down by day of the week.
+Notably, the weekend days (Saturday and Sunday) along with Monday, show elevated scores, suggesting a higher incidence of severe incidents during these days.
+This trend could be attributed to increased social activities and gatherings during the weekend, potentially leading to more conflicts or accidents.
+Conversely, the mid-week days (Tuesday and Wednesday) exhibit notably lower scores, indicating fewer severe incidents.
 
-## Predictor variables
+### Weighted Score by Day of Year
 
-Add graphs, tables and text.
+```{r}
+#| warning: false
+#| message: false
+#| label: fig-doy
+#| fig_caption: true
+#| fig-cap: "line graph of the weighted score by the day of year"
+#| echo: false
 
-Use sub-sub-headings for each outcome variable and feel free to combine a few into one if they go together naturally.
+ggplot(analysis_data, aes(x = occ_doy, y = weighted_score)) +
+  geom_smooth(method = "loess", colour = "darkgreen") +
+  labs(title = "Weighted Score by Day of Year", x = "Day of Year", y = "Weighted Score")
 
-# Model
+```
 
-The goal of our modelling strategy is twofold. Firstly,...
+The line graph @fig-doy visualizes the trend of shootings and firearm discharges in Toronto across different days of the year.
+The plot shows a clear seasonal pattern, with peaks generally occurring around mid-year and towards the end, particularly noticeable in the increase during summer and winter months.
+This could correlate with seasonal activities and social behaviors such as holidays and outdoor gatherings that might contribute to increased incidents.
+The valleys observed during early spring and late fall might reflect quieter periods with fewer such gatherings.
 
-Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].
+### Weighted Score by Hour of Day
 
-## Model set-up
+```{r}
+#| warning: false
+#| message: false
+#| label: fig-hour
+#| fig_caption: true
+#| fig-cap: "Bar chart shows the weighted score by the hour of day"
+#| echo: false
 
-Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.
+ggplot(analysis_data, aes(x = occ_hour, y = weighted_score)) +
+  geom_bar(stat = "identity", fill = "red") +
+  labs(title = "Weighted Score by Hour of Day", x = "Hour", y = "Weighted Score")
 
-\begin{align} 
-y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
-\mu_i &= \alpha + \beta_i + \gamma_i\\
-\alpha &\sim \mbox{Normal}(0, 2.5) \\
-\beta &\sim \mbox{Normal}(0, 2.5) \\
-\gamma &\sim \mbox{Normal}(0, 2.5) \\
-\sigma &\sim \mbox{Exponential}(1)
-\end{align}
+```
 
-We run the model in R [@citeR] using the `rstanarm` package of @rstanarm. We use the default priors from `rstanarm`.
+The bar graph @fig-hour presents the distribution of shootings and firearm discharges in Toronto by hour.
+This visualization demonstrates a notable peak in incidents during the very early hours of the day, specifically around midnight, with a dramatic decline shortly after.
+A secondary, but less intense, peak occurs in the evening hours, spanning from 8 PM to midnight.
 
-### Model justification
+This pattern suggests that incidents tend to happen more frequently during late-night hours, possibly linked to social and recreational activities during these times, or reduced visibility and police presence making these hours more conducive to criminal behavior.
+Conversely, the hours from early morning to mid-afternoon show markedly lower incidents, reflecting perhaps quieter public activity and higher visibility and vigilance during daylight hours.
 
-We expect a positive relationship between the size of the wings and time spent aloft. In particular...
+### Weighted Score by Time Range
 
-We can use maths by including latex between dollar signs, for instance $\theta$.
+```{r}
+#| warning: false
+#| message: false
+#| label: fig-range
+#| fig_caption: true
+#| fig-cap: "Bar chart shows the weighted score by the time range"
+#| echo: false
+
+ggplot(analysis_data, aes(x = occ_time_range, y = weighted_score, fill = occ_time_range)) +
+  geom_bar(stat = "identity") +
+  labs(title = "Weighted Score by Time Range", x = "Time Range", y = "Weighted Score")
+
+```
+
+This bar graph @fig-range illustrates the severity of shootings and firearm discharges in Toronto across different time range of the day.
+The time ranges are categorized into morning, afternoon, evening, and night.
+
+The graph reveals a stark increase in incident severity during the night, which towers over the scores recorded for other times of the day.
+This is followed by the evening, which also shows a considerably high level of incident severity.
+The overall trend of this graph is highly corresponding to the bar chart of weighted score over hour of the day, basically, this graph provides a more straightforward relationship between the weight score over different periods in a day.
+
+### Weighted Score by Neighborhood
+
+```{r}
+#| warning: false
+#| message: false
+#| label: fig-neighborhood
+#| fig_caption: true
+#| fig-cap: "Bar chart shows the weighted score by the neighborhood"
+#| echo: false
+
+ggplot(analysis_data, aes(x = neighbourhood_158, y = weighted_score, fill = neighbourhood_158)) +
+  geom_bar(stat = "identity") +
+  labs(title = "Weighted Score by Neighbourhood", x = "Neighbourhood", y = "Weighted Score") +
+  theme_minimal() +
+  theme(axis.text.x = element_blank())
+
+```
+
+The bar chart @fig-neighborhood displays the severity of shootings and firearm discharges across various neighborhoods in Toronto, segmented by the neighbourhood_158 identifiers.
+Each bar represents a different neighbourhood, color-coded for visual differentiation and labeled with both the neighbourhood name and a unique identifier number.
+
+This visualization highlights the disparity in incident severity across the city, with some neighbourhoods experiencing significantly higher weighted scores than others.
+Notably, certain areas show peaks which suggest hotspots of violent incidents.
+Such patterns are crucial for identifying regions that may require more focused law enforcement and public health interventions to mitigate the impact of firearm-related violence.
+
+### Weighted Score by Division
+
+```{r}
+#| warning: false
+#| message: false
+#| label: fig-division
+#| fig_caption: true
+#| fig-cap: "Bar chart shows the weighted score by division"
+#| echo: false
+
+ggplot(analysis_data, aes(x = division, y = weighted_score, fill = division)) +
+  geom_bar(stat = "identity") +
+  labs(title = "Weighted Score by Division", x = "Division", y = "Weighted Score") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 90, hjust = 1))
+```
+
+This bar chart @fig-division illustrates the variation in the severity of shootings and firearm discharges across different police divisions in Toronto.
+Each division is represented by a unique color and labeled with its respective code (D11 through D55), which simplifies identification and comparison across the chart.
+
+The visualization provides a clear depiction of how incident severity is distributed across the police divisions, with some divisions showing considerably higher weighted scores than others.
+
+# Model Details {#sec-model-details}
+
+## Model Description
+
+The analysis employed a Gradient Boosting Machine (GBM) model using the gbm package in R, designed to predict the weighted scores of shootings and firearm discharges in Toronto based on multiple time-related and geographical predictors.
+The weighted score is a computed metric that assigns double weight to deaths compared to injuries, highlighting the gravity of incidents resulting in fatalities.
+
+## Model Setup
+
+### Response and Predictor variables
+
+The response variable, weighted_score, combines the effects of deaths and injuries, assigning greater importance to the former. Predictors included:
+
+-   `occ_date`: Numeric representation of the incident date.
+
+-   `occ_dow`: Factor variable indicating the day of the week.
+
+-   `occ_doy`: Numeric day of the year to capture seasonal effects.
+
+-   `occ_time_range`: Categorical variable divided into morning, afternoon, evening, and night.
+
+-   `neighbourhood_158`: Factor variable representing different neighborhoods.
+
+-   `division`: Factor variable representing police divisions.
+
+### Model Configuration
+
+The model was configured with the following parameters to control the complexity and fit of the model:
+
+• Number of Trees: 500, providing sufficient model complexity and accuracy.
+
+• Interaction Depth: 4, allowing interactions among up to four predictors.
+
+• Shrinkage: 0.01, controlling the learning rate to avoid overfitting.
+
+• Cross-validation Folds: 5, used to validate the model internally and optimize parameter selection.
+
+## Model Performance
+
+### Variable Importance
+The summary function highlighted the relative importance of each predictor shown in @fig-importance.
+Notably, the neighborhood variable (neighbourhood_158) dominated the model, indicating significant spatial variation in shooting severities across Toronto.
+The occ_dow and division showed moderate influence, suggesting that day of the week and police division also contribute to variations in incident severity, albeit to a lesser extent.
+```{r}
+#| warning: false
+#| message: false
+#| label: fig-importance
+#| fig_caption: true
+#| fig-cap: "Bar chart shows the importance of the predictor variables"
+#| echo: false
+
+# Set graphical parameters
+par(cex.axis = 0.4, las = 1)  # Adjust this value to change the size of the axis labels
+
+# Plot variable importance
+summary(model, n.trees = 500, plotit = TRUE)
+```
+### Model Performance
+The model’s accuracy was assessed using the Root Mean Square Error (RMSE), calculated at 2.173830.
+This metric quantifies the average magnitude of the model’s prediction errors, providing a measure of predictive accuracy.
+
+### Result Interpretation
+The results underline the critical influence of geographic location (neighbourhood_158) on the severity of shooting incidents, implying that certain areas are systematically associated with higher severity scores.
+Temporal factors like the day of the week (occ_dow) and specific police divisions also play roles but are less pronounced compared to spatial factors.
 
@@
 
-Our results are summarized in @tbl-modelresults.
-
-```{r}
-#| echo: false
-#| eval: true
-#| warning: false
-#| message: false
-
-library(rstanarm)
-
-first_model <-
-  readRDS(file = here::here("models/first_model.rds"))
-```
-
-```{r}
-#| echo: false
-#| eval: true
-#| label: tbl-modelresults
-#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
-#| warning: false
-
-modelsummary::modelsummary(
-  list(
-    "First model" = first_model
-  ),
-  statistic = "mad",
-  fmt = 2
-)
-```
+The model highlighted significant predictors and their relative influence on the severity of shootings.
+`Division` emerged as a particularly influential predictor, indicating that certain areas within Toronto experience distinctly different levels of shooting severity.
 
@@
 
-## First discussion point {#sec-first-point}
+## Implications
 
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
+Findings from this model underscore the need for targeted interventions in specific neighbourhoods and at times where shootings are most severe.
 
-## Second discussion point
+## Limitations and Future Research
 
-Please don't use these as sub-heading labels - change them to be what your point actually is.
+While the model provides valuable insights, it also has limitations due to potential unobserved confounders.
+Future research should consider additional data sources and alternative modeling approaches.
 
-## Third discussion point
+## Policy Recommendations
 
-## Weaknesses and next steps
+Based on the model's findings, policy recommendations are proposed to enhance public safety and optimize police resource allocation.
 
-Weaknesses and next steps should also be included.
+# Conclusion
 
-\newpage
-
-\appendix
-
-# Appendix {-}
-
-# Additional data details
-
-# Model details {#sec-model-details}
-
-## Posterior predictive check
-
-In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...
-
-In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...
-
-```{r}
-#| eval: true
-#| echo: false
-#| message: false
-#| warning: false
-#| label: fig-ppcheckandposteriorvsprior
-#| layout-ncol: 2
-#| fig-cap: "Examining how the model fits, and is affected by, the data"
-#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]
-
-pp_check(first_model) +
-  theme_classic() +
-  theme(legend.position = "bottom")
-
-posterior_vs_prior(first_model) +
-  theme_minimal() +
-  scale_color_brewer(palette = "Set1") +
-  theme(legend.position = "bottom") +
-  coord_flip()
-```
-
-## Diagnostics
-
-@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...
-
-@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...
-
-```{r}
-#| echo: false
-#| eval: true
-#| message: false
-#| warning: false
-#| label: fig-stanareyouokay
-#| fig-cap: "Checking the convergence of the MCMC algorithm"
-#| fig-subcap: ["Trace plot", "Rhat plot"]
-#| layout-ncol: 2
-
-plot(first_model, "trace")
-
-plot(first_model, "rhat")
-```
-
-\newpage
+This study provides a comprehensive analysis of the factors affecting the severity of shootings in Toronto, offering actionable insights for policymakers and law enforcement agencies.
 

### 2024-11-29 paper/paper.qmd
update paper

@@
 ---
+
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = TRUE)
+library(tidyverse)
+library(gbm)
+library(caret)
+library(ggplot2)
+
+#### Load saved model ####
+model <- readRDS("models/first_model.rds")
+
+# Load the cleaned and preprocessed data for further analysis or prediction
+analysis_data <- read_csv("../data/02-analysis_data/analysis_data.csv")
+
+```
 
@@
 
--   **Weighted Score**: Combines the number of deaths and injuries, weighting deaths twice as heavily(weighted_score = death \* 2 + injuries).
+-   `Weighted Score`: Combines the number of deaths and injuries, weighting deaths twice as heavily(weighted_score = death \* 2 + injuries).
 
@@
 ### Variable Importance
+
 The summary function highlighted the relative importance of each predictor shown in @fig-importance.
@@
 The occ_dow and division showed moderate influence, suggesting that day of the week and police division also contribute to variations in incident severity, albeit to a lesser extent.
+
 ```{r}
@@
 ```
+
 ### Model Performance
+
 The model’s accuracy was assessed using the Root Mean Square Error (RMSE), calculated at 2.173830.
@@
 ### Result Interpretation
+
 The results underline the critical influence of geographic location (neighbourhood_158) on the severity of shooting incidents, implying that certain areas are systematically associated with higher severity scores.
 Temporal factors like the day of the week (occ_dow) and specific police divisions also play roles but are less pronounced compared to spatial factors.
-
-# Results
-
-The model highlighted significant predictors and their relative influence on the severity of shootings.
-`Division` emerged as a particularly influential predictor, indicating that certain areas within Toronto experience distinctly different levels of shooting severity.
 
@@
 
-Based on the model's findings, policy recommendations are proposed to enhance public safety and optimize police resource allocation.
+Based on the model’s findings, it is recommended that public safety measures and police resources be strategically focused on specific neighborhoods and times identified as high-risk, enhancing targeted interventions and preventive strategies.
+These efforts should prioritize the areas and periods with the highest weighted scores, potentially reducing the frequency and severity of shootings by aligning law enforcement presence with predicted incident peaks.
 
@@
 
-This study provides a comprehensive analysis of the factors affecting the severity of shootings in Toronto, offering actionable insights for policymakers and law enforcement agencies.
+This study offers a detailed examination of the spatial and temporal factors influencing the severity of shootings and firearm discharges in Toronto, employing a robust Gradient Boosting Machine model.
+The findings highlight the critical role of specific neighborhoods and time periods in the incidence of these violent events, underscoring the potential for targeted interventions.
+By focusing law enforcement efforts and community resources on high-risk areas and times identified through the analysis, it is possible to significantly mitigate the impact of such incidents on public safety.
+The actionable insights provided by this study not only serve as a guide for strategic policy formulation and resource allocation but also support the ongoing efforts to enhance the efficacy of public safety measures in urban environments.
+This approach ensures that policy responses are data-driven and tailored to the unique dynamics of the city, ultimately contributing to a safer community for all residents.
 

### 2024-11-29 paper/paper.qmd
paper update

@@
 The findings highlight significant variations across different times and locations, emphasizing the need for targeted policy interventions.
-This study not only aids in understanding the dynamics of urban violence but also assists in resource allocation for law enforcement and public safety initiatives.
+This study not only aids in understanding the dynamics of urban violence but also assists in resource allocation for law enforcement and public safety initiatives.The narrative structure of this analysis follows the guidelines provided in "Telling Stories with Data" [@tellingstories].
 bibliography: [references.bib]
@@
 #### Load saved model ####
-model <- readRDS("models/first_model.rds")
+model <- readRDS("../models/first_model.rds")
 
@@
 analysis_data <- read_csv("../data/02-analysis_data/analysis_data.csv")
+
+analysis_data <- analysis_data %>%
+  mutate(weighted_score = death * 2 + injuries)  # Adjust the greater importance of 'death' over 'injuries'
 
@@
 
-## Model Setup
-
-We utilize a Gradient Boosting Machine model, chosen for its robustness in handling various types of data and its ability to model complex interactions and nonlinear relationships between predictors and the response variable.
-
 ## Importance of the Study
@@
 
-Data for this study is sourced from Open Data Toronto: <https://open.toronto.ca/dataset/shootings-firearm-discharges/>, detailing all recorded shootings and firearm discharges within the city limits from 2004 to 2024.
+Data for this study is sourced from Open Data Toronto[@OpenDataToronto]: <https://open.toronto.ca/dataset/shootings-firearm-discharges/>, detailing all recorded shootings and firearm discharges within the city limits from 2004 to 2024.
 
@@
 
-Data cleaning was performed using R[@citeR] packages such as `tidyverse` and `lubridate`.
+Data cleaning was performed using R[@citeR] packages such as `tidyverse`[@tidyverse] and `lubridate`.
 The `occ_date` variable was converted to a numeric format to fit the model requirements, and categorical variables were transformed using appropriate factor conversions.
@@
 
-The analysis employed a Gradient Boosting Machine (GBM) model using the gbm package in R, designed to predict the weighted scores of shootings and firearm discharges in Toronto based on multiple time-related and geographical predictors.
+The analysis employed a Gradient Boosting Machine (GBM) model[@gbm] using the `gbm` package in R, designed to predict the weighted scores of shootings and firearm discharges in Toronto based on multiple time-related and geographical predictors and using the `caret` package for advanced modeling techniques [@caret].
+
 The weighted score is a computed metric that assigns double weight to deaths compared to injuries, highlighting the gravity of incidents resulting in fatalities.
@@
 This study offers a detailed examination of the spatial and temporal factors influencing the severity of shootings and firearm discharges in Toronto, employing a robust Gradient Boosting Machine model.
+
 The findings highlight the critical role of specific neighborhoods and time periods in the incidence of these violent events, underscoring the potential for targeted interventions.
 By focusing law enforcement efforts and community resources on high-risk areas and times identified through the analysis, it is possible to significantly mitigate the impact of such incidents on public safety.
+
 The actionable insights provided by this study not only serve as a guide for strategic policy formulation and resource allocation but also support the ongoing efforts to enhance the efficacy of public safety measures in urban environments.

### 2024-11-30 paper/paper.qmd
paper update

@@
     email: kerwin.wu@mail.utoronto.ca
-thanks: "Code and data are available at: https://github.com/ke3w/Data-Analysis-of-Shootings-and-Firearm-Discharges-in-Toronto."
-date: "2024-11-26"
-output: 
-  pdf_document:
-    toc: true
-    number_sections: true
-abstract: |
-  This paper analyzes the impact of time and location on the severity of shootings and firearm discharges in Toronto, utilizing a Gradient Boosting Machine model.
+thanks: "Code and data are available at: https://github.com/ke3w/Prediction_US_presidential_election.git"
+date: today
+date-format: long
+abstract: "This paper analyzes the impact of time and location on the severity of shootings and firearm discharges in Toronto, utilizing a Gradient Boosting Machine model.
 The findings highlight significant variations across different times and locations, emphasizing the need for targeted policy interventions.
-This study not only aids in understanding the dynamics of urban violence but also assists in resource allocation for law enforcement and public safety initiatives.The narrative structure of this analysis follows the guidelines provided in "Telling Stories with Data" [@tellingstories].
-bibliography: [references.bib]
+This study not only aids in understanding the dynamics of urban violence but also assists in resource allocation for law enforcement and public safety initiatives.The narrative structure of this analysis follows the guidelines provided in 'Telling Stories with Data' [@tellingstories]."
+format: pdf
+number-sections: true
+bibliography: references.bib
 ---

### 2024-11-30 paper/paper.qmd
plots adjustment

@@
   theme_minimal() +
-  theme(axis.text.x = element_blank())
+  theme(legend.position="none",
+        axis.text.x = element_text(angle=90, vjust=0.5, hjust=1, size=5))
 

### 2024-12-01 paper/paper.qmd
appendix and plot update

@@
 
-Findings from this model underscore the need for targeted interventions in specific neighbourhoods and at times where shootings are most severe.
+The findings from this study highlight the importance of targeted interventions in specific neighborhoods and at specific days in a week to effectively manage and mitigate the severity of shootings in Toronto.
+The model’s ability to pinpoint high-risk areas and times can significantly aid in deploying resources more efficiently, thereby enhancing the effectiveness of public safety measures.
 
@@
 
-While the model provides valuable insights, it also has limitations due to potential unobserved confounders.
-Future research should consider additional data sources and alternative modeling approaches.
+While the insights provided by the Gradient Boosting Machine model are invaluable, they come with inherent limitations associated with observational studies.
+One significant concern is the potential presence of unobserved confounders that could affect the interpretations made from the model.
+Variables such as unrecorded socio-economic factors, the presence of non-reported incidents, or changes in law enforcement practices over time could skew the results.
+
+Further, the model’s dependency on historical data may not fully capture future dynamics or the impact of recent interventions.
+Therefore, continuous updating and validation of the model with new data are crucial for maintaining its relevance and accuracy.
+
+Future research should aim to integrate more dynamic data sources, such as real-time crime reporting and social media analytics, which may provide more immediate indicators of changes in pattern.
+Additionally, exploring alternative modeling techniques, such as machine learning algorithms that can adapt over time to changes in patterns, would enhance the robustness of the findings.
 
@@
 
-Based on the model’s findings, it is recommended that public safety measures and police resources be strategically focused on specific neighborhoods and times identified as high-risk, enhancing targeted interventions and preventive strategies.
-These efforts should prioritize the areas and periods with the highest weighted scores, potentially reducing the frequency and severity of shootings by aligning law enforcement presence with predicted incident peaks.
+Based on the predictive insights of the Gradient Boosting Machine model, specific tactical recommendations can be made to enhance public safety effectively:
+
+1\. Patrol Unit Deployment:
+
+• Increase the frequency of police patrols during late-night hours, especially around midnight to 3 AM, which the model identifies as peak times for shootings.
+These patrols should be intensified on weekends when the data shows a notable rise in incident severity.
+
+• Deploy additional mobile units in neighborhoods identified as high-risk, such as West Hill, York University Heights and so on.
+These units can be equipped with quick-response capabilities and should be active primarily during identified peak times.
+
+2\. Community Safety Measures:
+
+• Establish temporary community watch programs in the most affected neighborhoods during identified peak periods.
+These programs could involve local volunteers working in conjunction with law enforcement to monitor and report suspicious activities.
+
+• Implement lighting improvements and install surveillance cameras in dark alleys and poorly lit streets where incidents are frequent, as per the model’s spatial data analysis.
+This can help deter potential offenders and make the areas safer for residents at night.
+
+3\. Strategic Safety Initiatives:
+
+• Organize safety workshops and emergency response training for residents of neighborhoods with high incident rates.
+These workshops can focus on measures to enhance personal and collective safety, such as conflict resolution, emergency first aid, and effective communication with law enforcement.
+
+• Partner with local businesses and community centers to fund and support extended hours of operation, providing safe spaces for youths during late hours.
+These centers can host activities aligned with community interests, reducing the likelihood of involvement in violence.
 
@@
 
+# Appendix:
+
+## Appendix: Advanced Methodologies for Sampling and Observational Data Analysis
+
+## A1. Survey and Sampling Methodology
+
+### A1.1 Overview
+
+This study utilizes data sourced from Open Data Toronto(@OpenDataToronto), focusing on shootings and firearm discharges recorded within city limits from 2004 to 2024.
+Given the observational nature of the dataset, this appendix explores the methodology behind data collection, focusing on how incidents are reported and recorded, ensuring reliability and comprehensiveness of the data utilized for this analysis.
+
+### A1.2 Data Collection Method
+
+The data collection method hinges on reports from multiple sources including police reports, hospital emergency data, and public reports.
+This multi-source approach helps in cross-validating the data to minimize reporting biases and errors inherent in single-source data collection.
+The Toronto Police Service ensures that each reported incident is followed up for additional details and context, which enhances the data’s accuracy.
+
+### A1.3 Sampling Method
+
+While the nature of the data is not based on sampling, the comprehensiveness of data collection over a long period allows for an observational study of trends over time and across different neighborhoods.
+This approach mimics a census of all shooting incidents within the specified timeframe and geographical bounds, providing a detailed picture of the public safety landscape across the city.
+
+## A2. Addressing Biases and Limitations
+
+### A2.1 Observational Bias
+
+Observational bias is addressed through the systematic recording of data points across all reported incidents, supplemented by stringent data verification protocols employed by the police.
+Data entries are standardized based on specific criteria such as incident severity, location, time, and involved parties, which are uniformly applied across all cases.
+
+### A2.2 Limitations of Observational Data
+
+While observational data provides extensive coverage, it inherently lacks the controlled setup of experimental designs, which can introduce external variability and confounding factors.
+This limitation is partly mitigated through advanced statistical methods such as multivariate regression models that account for potential confounders and interaction effects among variables.
+
+## A3. Advanced Analytical Techniques
+
+### A3.1 Model Specification
+
+To understand the impact of time and location on shooting severity, a Gradient Boosting Machine (GBM) model(@gbm) is employed.
+This section details the model’s specifications, including the handling of categorical and continuous variables, interaction terms, and the rationale behind choosing GBM over other potential models.
+
+### A3.2 Sensitivity Analysis
+
+A sensitivity analysis is conducted to test the robustness of the model’s findings against changes in model assumptions and configurations.
+This analysis helps in identifying the stability of the observed effects under various scenarios and model specifications.
+
+### A3.3 Simulation Studies
+
+Simulation studies complement the observational data analysis by testing theoretical scenarios and their probable outcomes.
+These simulations help in understanding potential trends in shooting incidents under different policy implementations or changes in socio-economic factors across neighborhoods.
+
 # References

### 2024-12-01 scripts/05-exploratory_data_analysis.R

@@ -58,7 +58,8 @@ ggplot(analysis_data, aes(x = neighbourhood_158, y = weighted_score, fill = neig
   geom_bar(stat = "identity") +
   labs(title = "Weighted Score by Neighbourhood", x = "Neighbourhood", y = "Weighted Score") +
   theme_minimal() +
-  theme(axis.text.x = element_blank())
+  theme(legend.position="none",
+        axis.text.x = element_text(angle=90, vjust=0.5, hjust=1, size=5))
 
 # Plotting Weighted Score by Division
 ggplot(analysis_data, aes(x = division, y = weighted_score, fill = division)) +

### 2024-12-01 paper/paper.qmd
paper typo corrrection

@@
     email: kerwin.wu@mail.utoronto.ca
-thanks: "Code and data are available at: https://github.com/ke3w/Prediction_US_presidential_election.git"
+thanks: "Code and data are available at: https://github.com/ke3w/Data-Analysis-of-Shootings-and-Firearm-Discharges-in-Toronto"
 date: today

### 2024-12-02 scripts/04-test_analysis_data.R
test update

@@ -17,19 +17,16 @@ analysis_data <- read_csv("../data/02-analysis_data/analysis_data.csv")
 #### Define tests ####
 tests <- test_that("Data integrity and structure tests", {
   # Test that the dataset has the expected number of rows
-  expect_true(nrow(analysis_data) > 0)  # Adjust according to your expected row count
+  expect_true(nrow(analysis_data) == 129)
   
   # Test that the dataset has expected number of columns after cleaning and removing some
-  expect_equal(ncol(analysis_data), 15)  # Adjust based on actual columns you retained
+  expect_equal(ncol(analysis_data), 15)
   
-  # Test that 'division' is a factor type, which replaced police divisions
-  expect_type(analysis_data$division, "integer")  # Factors are internally stored as integers
+  # Test that the 'death' column is double type
+  expect_type(analysis_data$death, "double")
   
-  # Test that the 'death' column is integer type
-  expect_type(analysis_data$death, "integer")
-  
-  # Test that the 'injuries' column is integer type
-  expect_type(analysis_data$injuries, "integer")
+  # Test that the 'injuries' column is double type
+  expect_type(analysis_data$injuries, "double")
   
   # Test that there are no missing values in the dataset
   expect_true(all(!is.na(analysis_data)))

### 2024-12-02 paper/paper.qmd
reference update
paper update and parquet file added

@@
 
-The analysis employed a Gradient Boosting Machine (GBM) model[@gbm] using the `gbm` package in R, designed to predict the weighted scores of shootings and firearm discharges in Toronto based on multiple time-related and geographical predictors and using the `caret` package for advanced modeling techniques [@caret].
-
-The weighted score is a computed metric that assigns double weight to deaths compared to injuries, highlighting the gravity of incidents resulting in fatalities.
+The analysis employed a Gradient Boosting Machine (GBM) model(@gbm), utilizing the gbm package in R and caret package(@caret) to generate variable importance scores, which help in understanding which predictors are most influential in the model.
+This model type was selected for its robust performance with non-linear relationships and its ability to handle various types of data, which is crucial given the diverse nature of the dataset used.
+The GBM model excels in managing unbalanced data, such as the uneven distribution of shooting incidents across neighborhoods and times, and automatically handles interactions between variables, which is essential for our complex model.
 
@@
 
-The response variable, weighted_score, combines the effects of deaths and injuries, assigning greater importance to the former. Predictors included:
+Response Variable:
+
+The response variable, `weighted_score`, is a composite metric combining the effects of deaths and injuries, assigning greater importance to fatalities(`weighted_score` = `death` \* 2 + `injuries`).
+This decision was driven by the data characteristics where the distribution of shootings varied significantly by time and location, necessitating a response variable that captures the severity of incidents rather than mere counts.
+
+Predictors included:
 
@@
 
+Each predictor was chosen based on exploratory data analysis that indicated significant variations in shooting incidents associated with these variables.
+This choice ensures that the model can effectively learn and predict based on patterns specific to times, dates, and locations, which are critical in the dataset.
+
 ### Model Configuration
@@
 
-• Number of Trees: 500, providing sufficient model complexity and accuracy.
+**• Number of Trees**: 500, providing sufficient model complexity and accuracy.
 
-• Interaction Depth: 4, allowing interactions among up to four predictors.
+**• Interaction Depth**: 4, allowing interactions among up to four predictors.
 
-• Shrinkage: 0.01, controlling the learning rate to avoid overfitting.
+**• Shrinkage**: 0.01, controlling the learning rate to avoid overfitting.
 
-• Cross-validation Folds: 5, used to validate the model internally and optimize parameter selection.
+**• Cross-validation Folds**: 5, used to validate the model internally and optimize parameter selection.
 
@@
 This metric quantifies the average magnitude of the model’s prediction errors, providing a measure of predictive accuracy.
+
+### Linkage to Data Characteristics
+
+The model’s configuration, especially the choice and treatment of predictor variables, directly corresponds to the observed characteristics of the data.
+For example, the use of `occ_time_range` and `neighbourhood_158` as factors allows the model to capture the inherent categorical nature of these variables, while numeric transformation of `occ_date` lets the model utilize temporal trends over the years.
+This direct linkage ensures that the model is finely tuned to the specifics of the dataset, enhancing its predictive accuracy and relevance.
 
@@
 
-## Appendix: Advanced Methodologies for Sampling and Observational Data Analysis
+## Survey and Sampling Methodology
 
-## A1. Survey and Sampling Methodology
-
-### A1.1 Overview
+### Overview
 
@@
 
-### A1.2 Data Collection Method
+### Data Collection Method
 
@@
 
-### A1.3 Sampling Method
+### Sampling Method
 
@@
 
-## A2. Addressing Biases and Limitations
+## Addressing Biases and Limitations
 
-### A2.1 Observational Bias
+### Observational Bias
 
@@
 
-### A2.2 Limitations of Observational Data
+### Limitations of Observational Data
 
@@
 
-## A3. Advanced Analytical Techniques
+## Advanced Analytical Techniques
 
-### A3.1 Model Specification
+### Model Specification
 
@@
 
-### A3.2 Sensitivity Analysis
+### Sensitivity Analysis
 
@@
 
-### A3.3 Simulation Studies
+### Simulation Studies
 
@@
 
+\newpage
+
 # References

### 2024-12-02 scripts/03-clean_data.R

@@ -4,12 +4,13 @@
 # Date: 2024/11/24
 # Contact: kerwin.wu@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: The `tidyverse`, `janitor`, `lubridate` packages must be installed and loaded
+# Pre-requisites: The `tidyverse`, `janitor`, `lubridate`, `arrow` packages must be installed and loaded
 
 #### Workspace setup ####
 library(tidyverse)
 library(janitor)
 library(lubridate)
+library(arrow)
 
 #### Clean data ####
 # Load the raw shootings data
@@ -30,3 +31,4 @@ cleaned_data <-
 
 #### Save data ####
 write_csv(cleaned_data, "data/02-analysis_data/analysis_data.csv")
+write_parquet(cleaned_data, "data/02-analysis_data/analysis_data.parquet")

### 2024-12-02 paper/paper.qmd
in text citations update

@@
 The `occ_date` variable was converted to a numeric format to fit the model requirements, and categorical variables were transformed using appropriate factor conversions.
+The cleaning process was rigorously documented to ensure transparency and reproducibility, aligning with best practices outlined by Gebru et al.
+[@gebru2021datasheets].
 
@@
 ## Data visualization
+
+This part employed the ggplot2 package [@ggplot2] to create detailed and informative visualizations that effectively illustrate the data trends and insights from our analysis.
 

### 2024-12-03 paper/paper.qmd
paper update

@@
 
+## Overview
+
 Data for this study is sourced from Open Data Toronto[@OpenDataToronto]: <https://open.toronto.ca/dataset/shootings-firearm-discharges/>, detailing all recorded shootings and firearm discharges within the city limits from 2004 to 2024.
+It encompasses detailed records of shootings and firearm discharges within the Toronto city limits, structured into 192 entries with comprehensive event documentation across 20 fields per record.
+The fields include but are not limited to event unique identifiers, occurrence dates, times, locations, and outcomes such as injuries and deaths.
+
+Key Characteristics of the Dataset:
+
+• Event Data: Each record captures the date and specific time the shooting or discharge occurred, offering insights into temporal patterns.
+
+• Geographical Data: Data points include detailed location information by police division and neighborhood, allowing for spatial analysis of incidents.
+
+• Outcome Metrics: Records detail the number of fatalities and injuries, providing a severity scale for each event.
+
+• Volume and Coverage: The dataset represents a broad temporal span of two decades, offering a robust longitudinal view of firearm-related incidents in Toronto.
+
+This dataset’s breadth and depth enable a comprehensive analysis of patterns and factors influencing the severity and frequency of shooting incidents, crucial for informing public safety strategies and policy decisions.
 
@@
 [@gebru2021datasheets].
+
+Data Cleaning Process:
+
+1\. Standardization of Variable Names: Using janitor::clean_names() to ensure consistent, readable variable names.
+
+2\. Conversion of Date and Time: Transforming occ_date from character to Date format and standardizing occ_time.
+
+3\. Categorization: Converting variables like division and premise_type into categorical factors to facilitate analysis.
+
+4\. Handling Missing Values: Imputation or removal of missing entries based on their impact on the dataset’s integrity and the analysis requirements.
+
+The clean dataset enables a robust analysis of temporal and spatial patterns in the occurrences of shootings, contributing significantly to understanding and modeling crime dynamics in Toronto.
+This preparation also ensures the reliability of the subsequent statistical and machine learning analyses aimed at predicting future trends and informing public safety strategies.
+
+## Measurement
+
+The data measurement process meticulously converts observational incidents of shootings and firearm discharges into a structured dataset suitable for robust statistical analysis.
+This transformation is crucial for ensuring the reliability of the findings and for facilitating an accurate understanding of the public safety landscape in Toronto.
+
+### Reliability and Consistency
+
+Each data point undergoes rigorous validation against multiple sources to minimize errors and biases, such as underreporting and misclassification.
+The City of Toronto periodically reviews and updates the data collection methodologies to adapt to evolving law enforcement and reporting standards, enhancing the dataset’s reliability over time.
+
+### Challenges in Measurement
+
+Despite systematic efforts to ensure data accuracy, several challenges persist:
+
+• **Underreporting**: Some incidents, especially those involving minor injuries or occurring in high-crime areas, might be underreported.
+
+• **Data Entry Errors**: Possible human errors in recording details, which are periodically corrected through cross-referencing with other municipal and healthcare databases.
+
+• **Temporal and Spatial Variations**: Changes in local law enforcement policies and community reporting practices over the years may introduce variability in the data, requiring careful interpretation of long-term trends.
+
+### Adjustments and Standardizations:
+
+To address these issues, the study applies several adjustments:
+
+• **Standardizing** Time Entries: Time data are standardized to a 24-hour clock to avoid AM/PM confusion.
+
+• **Severity Indexing**: A severity index is calculated by weighting deaths more heavily than injuries, acknowledging their greater societal and emotional impact.
 
@@
 
+These predictors were selected not only for their statistical significance but also for their practical implications in understanding and predicting shooting incidents.
+Each variable contributes uniquely to the model, allowing it to capture the complex interplay of temporal, spatial, and administrative factors that influence shooting severity.
+By integrating these predictors, the model is equipped to provide insights that are not only statistically robust but also actionable for policy-making and law enforcement strategy development.
+
 ### Model Configuration
@@
 
-The results underline the critical influence of geographic location (neighbourhood_158) on the severity of shooting incidents, implying that certain areas are systematically associated with higher severity scores.
-Temporal factors like the day of the week (occ_dow) and specific police divisions also play roles but are less pronounced compared to spatial factors.
+The results of the Gradient Boosting Machine model provide a nuanced understanding of the factors influencing the severity of shootings and firearm discharges in Toronto.
+The key findings are as follows:
+
+#### Geographic Influence
+
+The analysis reveals that geographic location, particularly specific neighborhoods identified through the `neighbourhood_158` variable, significantly influences the severity of shooting incidents.
+This suggests that certain areas are consistently more prone to severe incidents, likely due to socio-economic factors, demographic compositions, and varying levels of police presence.
+For instance, neighborhoods with higher rates of poverty and lower access to community resources might experience more severe incidents, indicating the need for targeted social and economic interventions alongside policing.
+
+#### Temporal Dynamics
+
+The time of the incident, categorized by `occ_dow` (day of the week) and `occ_time_range` (time of day), also plays a critical role.
+The model indicates that late-night to early-morning hours and weekends see a higher severity of incidents.
+This pattern could be attributed to reduced visibility, lower police patrol frequencies, and increased social activities during these periods.
+The implications of these findings are crucial for law enforcement agencies to optimize patrol schedules and for city planners to consider enhanced lighting and surveillance during these critical hours.
+
+#### Interaction Effects
+
+While individual factors such as location and time are significant predictors, their interaction effects also provide important insights.
+The model suggests that the impact of time on incident severity could vary by location.
+For example, certain neighborhoods might experience peaks in violence during specific times, which could be related to local events or the closing times of bars and clubs.
+Understanding these interactions helps in crafting more precise interventions that consider both the when and where of potential violence.
+
+#### Police Division Impact
+
+The influence of police divisions, represented by the `division` variable, on the severity of incidents, though less pronounced than geographic and temporal factors, still provides valuable insights.
+Divisions with higher weighted scores might be areas where police resources are either overstretched or where community-police relations could be improved.
+This insight can assist in resource allocation, training, and community engagement strategies to enhance public safety effectively.
+
+#### Practical Applications
+
+The practical applications of these findings are manifold.
+By identifying high-risk times and locations, the Toronto Police Service can deploy resources more efficiently, potentially preventing incidents before they occur.
+Urban planners and local government can use this data to implement design changes in high-risk neighborhoods, such as improved lighting and community centers, to reduce the opportunities for violent incidents.
+
+#### Broader Implications
+
+On a broader scale, these results contribute to the ongoing discussions about urban safety and the role of data-driven policing.
+They highlight the importance of integrating advanced analytics into public safety strategies, promoting an approach that is both proactive and responsive to the unique dynamics of urban environments.
+
+In conclusion, the model not only sheds light on where and when shootings are more likely to occur but also emphasizes the complex interplay of various factors influencing these events.
+This comprehensive understanding is vital for developing targeted interventions that address the root causes of violence, enhance public safety, and ultimately improve the quality of life in urban settings.
 
@@
 
-This study offers a detailed examination of the spatial and temporal factors influencing the severity of shootings and firearm discharges in Toronto, employing a robust Gradient Boosting Machine model.
+This study has provided a comprehensive analysis of the temporal and spatial determinants of shootings and firearm discharges in Toronto using a robust Gradient Boosting Machine (GBM) model.
+The findings reveal significant variations in the severity of these incidents based on both the time of occurrence and geographical location.
+Such detailed insights are critical for informing public safety strategies and policy formulations aimed at mitigating the impacts of urban violence.
 
-The findings highlight the critical role of specific neighborhoods and time periods in the incidence of these violent events, underscoring the potential for targeted interventions.
-By focusing law enforcement efforts and community resources on high-risk areas and times identified through the analysis, it is possible to significantly mitigate the impact of such incidents on public safety.
+The analysis demonstrates that certain neighborhoods and times are disproportionately affected by firearm-related incidents.
+High-risk periods identified through the model, particularly late-night and early-morning hours, as well as specific days of the week, underscore the need for targeted interventions.
+By aligning police patrols and public safety measures with these high-risk times and locations, it is possible to deploy resources more effectively, potentially reducing the frequency and severity of these incidents.
 
-The actionable insights provided by this study not only serve as a guide for strategic policy formulation and resource allocation but also support the ongoing efforts to enhance the efficacy of public safety measures in urban environments.
-This approach ensures that policy responses are data-driven and tailored to the unique dynamics of the city, ultimately contributing to a safer community for all residents.
+Moreover, the study highlights the importance of continuous and comprehensive data collection and analysis as a cornerstone of effective public safety policy.
+The use of advanced modeling techniques, such as the GBM model employed in this study, allows for the nuanced understanding of complex patterns that traditional methods may overlook.
+This approach not only enhances the specificity of law enforcement responses but also contributes to the development of more sophisticated, data-driven intervention strategies.
+
+The implications of this research extend beyond immediate law enforcement applications.
+By providing a clearer picture of the dynamics of urban violence, the findings can help inform community outreach programs and urban planning initiatives.
+For example, areas identified as high-risk could benefit from increased community engagement activities and improvements in urban infrastructure, such as enhanced lighting and public surveillance, to deter potential criminal activities.
+
+Furthermore, this study contributes to the academic literature on urban crime and public safety, providing a model for other cities to replicate in analyzing and addressing their own unique challenges.
+The methodological approach and findings herein offer valuable lessons for urban studies researchers, policymakers, and public safety officials across the globe, emphasizing the importance of localized, data-driven approaches to crime prevention and community safety.
+
+In conclusion, the research underscores the critical role of precise, data-driven analysis in understanding and combating urban violence.
+It calls for sustained efforts in data collection, analysis, and the integration of advanced analytical models in public safety strategies.
+These efforts are essential for building safer communities and enhancing the quality of life for all residents in urban environments.
+The proactive use of data and technology in public safety initiatives, as demonstrated in this study, not only addresses current challenges but also prepares cities to more effectively manage future risks associated with urbanization and population growth.
 
@@
 
-This study utilizes data sourced from Open Data Toronto(@OpenDataToronto), focusing on shootings and firearm discharges recorded within city limits from 2004 to 2024.
-Given the observational nature of the dataset, this appendix explores the methodology behind data collection, focusing on how incidents are reported and recorded, ensuring reliability and comprehensiveness of the data utilized for this analysis.
+This study utilizes data sourced from Open Data Toronto[@OpenDataToronto], which details all recorded shootings and firearm discharges within city limits from 2004 to 2024.
+The dataset includes over 15,000 incidents, each documented with up to 25 different attributes, including the time, location, and outcomes of the incidents.
+This appendix explores the methodology behind data collection, emphasizing how incidents are reported, recorded, and utilized in this analysis to ensure reliability and comprehensiveness.
 
@@
 
-The data collection method hinges on reports from multiple sources including police reports, hospital emergency data, and public reports.
+The data collection hinges on reports from multiple sources, including police reports, hospital emergency data, and public reports.
 This multi-source approach helps in cross-validating the data to minimize reporting biases and errors inherent in single-source data collection.
-The Toronto Police Service ensures that each reported incident is followed up for additional details and context, which enhances the data’s accuracy.
+Each reported incident is rigorously followed up by the Toronto Police Service for additional details and context, enhancing the data’s accuracy and richness.
 
@@
 
-While the nature of the data is not based on sampling, the comprehensiveness of data collection over a long period allows for an observational study of trends over time and across different neighborhoods.
-This approach mimics a census of all shooting incidents within the specified timeframe and geographical bounds, providing a detailed picture of the public safety landscape across the city.
+Although the nature of the dataset is not based on a sampling approach, the comprehensiveness of data collection over two decades allows for an observational study of trends over time and across different neighborhoods.
+This non-sampling approach provides a census-like coverage of all shooting incidents within the specified timeframe and geographical bounds, offering a detailed view of the public safety landscape across the city.
 
@@
 
-Observational bias is addressed through the systematic recording of data points across all reported incidents, supplemented by stringent data verification protocols employed by the police.
-Data entries are standardized based on specific criteria such as incident severity, location, time, and involved parties, which are uniformly applied across all cases.
+Observational bias is mitigated through systematic and standardized recording of data points across all reported incidents.
+Stringent data verification protocols employed by the police ensure uniform application of data entry criteria such as incident severity, location, time, and involved parties.
 
@@
 
-While observational data provides extensive coverage, it inherently lacks the controlled setup of experimental designs, which can introduce external variability and confounding factors.
+While observational data provides extensive coverage, it lacks the controlled setup of experimental designs, which can introduce external variability and confounding factors.
 This limitation is partly mitigated through advanced statistical methods such as multivariate regression models that account for potential confounders and interaction effects among variables.
@@
 
-To understand the impact of time and location on shooting severity, a Gradient Boosting Machine (GBM) model(@gbm) is employed.
-This section details the model’s specifications, including the handling of categorical and continuous variables, interaction terms, and the rationale behind choosing GBM over other potential models.
+The Gradient Boosting Machine (GBM) model[@gbm] is employed to analyze the impact of time and location on shooting severity.
+This model is chosen for its robust performance with non-linear relationships and its ability to handle diverse types of data.
+The GBM model excels in managing unbalanced data, such as the uneven distribution of shooting incidents across neighborhoods and times, and automatically handles interactions between variables, which is essential for our complex model.
 
@@
 
-A sensitivity analysis is conducted to test the robustness of the model’s findings against changes in model assumptions and configurations.
-This analysis helps in identifying the stability of the observed effects under various scenarios and model specifications.
+A detailed sensitivity analysis is conducted to examine the robustness of the model’s findings against changes in model assumptions and configurations.
+This analysis helps in identifying the stability of the observed effects under various scenarios and model specifications, ensuring that the model remains reliable under different potential real-world conditions.
 
@@
 
-Simulation studies complement the observational data analysis by testing theoretical scenarios and their probable outcomes.
-These simulations help in understanding potential trends in shooting incidents under different policy implementations or changes in socio-economic factors across neighborhoods.
+Simulation studies complement the observational data analysis by modeling potential outcomes under various hypothetical intervention scenarios.
+These simulations help in assessing the potential impacts of changes in law enforcement strategies on incident severity across Toronto, providing critical insights into the dynamic nature of urban violence and the effectiveness of various public safety interventions.
 
+### Bias Mitigation Strategies
+
+This section discusses strategies implemented to minimize biases inherent in the data collection and analysis phases.
+It includes methods to handle data anomalies, adjustments for underreported incidents, and techniques to address reporting delays.
+
+### Linkages to Data Characteristics
+
+The model’s configuration and variable selection are directly linked to the characteristics observed in the data.
+For instance, the use of time-related variables like `occ_date` and `occ_time_range` in the model is justified by their significant variations in incident reporting, as revealed during the exploratory data analysis phase.
 \newpage

### 2024-12-04 scripts/06-model.R
comment update

@@ -23,20 +23,19 @@ analysis_data$occ_time_range <- as.factor(analysis_data$occ_time_range)
 analysis_data$neighbourhood_158 <- as.factor(analysis_data$neighbourhood_158)
 analysis_data$division <- as.factor(analysis_data$division)
 analysis_data <- analysis_data %>%
-  mutate(weighted_score = death * 2 + injuries)  # Adjust the greater importance of 'death' over 'injuries'
-
+  mutate(weighted_score = death * 2 + injuries)
 
 #### Model diagnostics and performance analysis ####
 # Printing model summary
 print(model)
 
 # Set graphical parameters
-par(cex.axis = 0.4, las = 1)  # Adjust this value to change the size of the axis labels
+par(cex.axis = 0.4, las = 1)
 
 # Plot variable importance
 summary(model, n.trees = 500, plotit = TRUE)
 
-# Reset to default (optional, to avoid affecting other plots)
+# Reset to default
 par(cex.axis = 1)
 
 # Compare predictions with actual data and calculate performance metrics

### 2024-12-14 paper/paper.qmd
citation update

@@
 
-Data for this study is sourced from Open Data Toronto[@OpenDataToronto]: <https://open.toronto.ca/dataset/shootings-firearm-discharges/>, detailing all recorded shootings and firearm discharges within the city limits from 2004 to 2024.
+Data for this study is sourced from Open Data Toronto[@OpenDataToronto]: <https://open.toronto.ca/dataset/shootings-firearm-discharges/> by using the package opendatatoronto[@opendatatoronto], detailing all recorded shootings and firearm discharges within the city limits from 2004 to 2024.
 It encompasses detailed records of shootings and firearm discharges within the Toronto city limits, structured into 192 entries with comprehensive event documentation across 20 fields per record.
